Mixture Models for Single Cell Assays

### 2012-02-04 MIMOSA.tex
modified the abstract and intro to be more generic, referring more to single cell assays in general.
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\begin{document}
\maketitle
\begin{abstract}
Flow cytometry, specifically the intracellular cytokine staining (ICS) assay is an important tool to assess immune response to vaccines in clinical trials, providing both phenotypic and functional information about the immune system. The goal of the ICS assay is to determine whether an individual's immune system responds to a vaccine, as well as the nature of the response. ICS assays test multiple antigens and multiple cytokines across hundreds of individuals, with the usual goal of identifying individuals in whom some rare subpopulation of cells is greater in a stimulated conditions than in a control. Maximizing sensitivity and specificity to detect responders to vaccination is one of the most important considerations in designing an ICS assay, particularly due to the rarity of the cell subpopulations being measured. The measured cell subpopulations are usually rare, such as antigen--specific cytokine producing T--cells, and are often measured as a fraction of their parent population. One assay typically measures multiple cytokines simultaneously. Consequently, ICS data is typically analyzed as a series of 2x2 contingency tables using Fisher's exact test. This is problematic for a number of reasons. Fisher's exact test can be overly conservative for small counts. Additionally, the data in the 2x2 table for the ICS assay does not truly have fixed margins since the stimulated and unstimulated cells come from independent experiments. In this paper we present a coherent empirical-Bayes mixture model framework that models cytokine–specific T-cell response across all individuals in a study simultaneously, while modelling the stimulated and unstimulated cell counts independently. We show that our model increases the sensitivity and specificity for positivity calls compared to the classical approach, using simulations and real data from vaccine trials.

\end{abstract}
\section{Introduction}
Intracellular cytokine staining assays are used in the context of vaccine trials to provide a sensitive and quantitative assessment of antigen--specific T cell response to vaccine, including phenotyping and measurement of multiple effector functions~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw}.

%An automated system for analyzing data from ICS assays has been developed by LabKey Software in collaboration with the HVTN, and is being used for current HIV vaccine trials~\cite{Shulman:2008jv}. A limitation of this system is that it relies on manual gating using external tools, and thus all the limitations of manual gating apply. The process is time consuming, error prone and subjective. Laboratories using the system overcome this drawback by performing highly controlled experiments that are suitable for use with template gates and requiring only occasional manual adjustment by the analyst~\cite{Horton:2007tsa,Shulman:2008jv}. Despite these mitigating factors, gating rare cytokine positive cell populations in ICS assays remains challenging, and biologically interesting markers with antibodies that have high staining variability cannot be analyzed in an automated fashion since they require considerable effort to make manual adjustments.
%
%In order to make further advancements in computational approaches to flow cytometry data analysis, computational researchers must be able to access the current ``gold--standard'' manual gating analysis to make quantitative comparisons. Accessibility of manual gating and data preprocessing pipelines for flow cytometry studies has been sporadic at best. LabKey allows access to population statistics extracted from flowJo workspaces, but doesn't allow access to the manual gates directly~\cite{Shulman:2008jv}. The BioConductor package \textit{flowFlowJo} allows import of a subset of older flowJo workspaces ($\leq$ v7.0) into BioConductor for use with \textit{flowCore} and other packages\cite{Hahne:2009vv, Gosink:2009iq}. However, this package does not support newer flowJo workspaces (>7.0) generated by the Macintosh version of the software, which are used by many laboratories, and limitations of the data structures representing an FCM workflow in BioConductor do not readily allow for sample to sample variability in the gate definitions for corresponding populations. Furthermore \textit{flowFlowJo} and the BioConductor flow tools are impractical for the analysis of realistically sized clinical flow cytometry data sets, which often exceed the memory capacity of the workstations running the software.

%To address problems gating cell subpopulations when there is substantial sample--to--sample variability algorithms have been developed for FCM data normalization, including \textit{fdaNorm} and \textit{gaussNorm} in the \textit{flowStats} package in BioConductor~\cite{Hahne:2010uh}. These algorithms are designed to normalize FCM data on a per--channel basis across multiple samples. Briefly, these normalization algorithms perform density estimation, feature extraction (peak identification), peak--matching (registration), and finally alignment. However, their utility has remained largely unproven in real--world applications, partly due to the inherent limitations already mentioned of working with large FCM data sets in BioConductor, partly due to difficulties with the peak--matching step when the data densities are multi--modal (i.e. three or more modes) and variable across samples, and partly due to difficulties in accessing real--world manually gated data for direct comparison. 
%
%These usability issues have hampered the adoption of the otherwise powerful and cutting--edge FCM data analysis toolset available in BioConductor. In this paper, we present a series of packages and extensions to the BioConductor FCM toolset that address these problems.

%The \textit{flowWorkspace} package allows the import of manual gates, compensation matrices and data transformations from flowJo workspaces generated by Macintosh versions of flowJo, making a majority of the FCM analyses archived in laboratories available for analysis in BioConductor. \textit{flowWorkspace} defines several new data structures for working with FCM data in BioConductor, including the \textit{GatingHierarchy} representing the hierarchical set of gates applied to an FCM file, and the \textit{GatingSet} representing a collection of \textit{GatingHierarchies} on multiple FCM files. Importantly, while the \textit{GatingSet} requires all samples to have the same subpopulations defined, it allows the precise gate definitions to vary across samples, more accurately reflecting the manner in which gates may be adjusted for individual samples within \textit{flowJo} by the data analyst. 
%
%The \textit{flowWorkspace} package is fully integrated with the new \textit{ncdfFlow} package that provides disk--based access to FCM data via the netCDF C library. The package defines a new \textit{ncdfFlowSet} object that inherits from flowCore's \textit{flowSet}. We have further extended the \textit{flowCore} package to ensure existing methods and algorithms are compatible with \textit{ncdfFlow}. Together, \textit{flowWorkspace} and \textit{ncdfFlow} allow  real--world FCM datasets many Gigabytes in size to be analyzed using the BioConductor toolset on regular workstations.
%
%Finally, we have made several improvements to the \textit{fdaNorm} FCM data normalization algorithm in \textit{flowStats}, making it more robust, and amenable to integration in an FCM data analysis pipeline. First, we have adapted the algorithm to take advantage of the \textit{ncdfFlow} package allowing it to work on large data sets. We have also added the ability to chunk the data into several parts, making the algorithm more memory efficient, and importantly, we have added the ability to normalize FCM data to a target distribution defined by a specific FCM file. Support for \textit{flowWorkspace} has been added to the normalization algorithm, allowing the user to easily include a normalization step for flowJo defined manual gates. Finally, and perhaps most importantly, we have developed a new sequential normalization strategy in the context of manual gating. Rather than normalizing FCM data once, at the beginning of gating, our approach performs a normalization step prior to each hierarchical gating step. This has several benefits: primarily, population matching is simplified for each normalization step since overlapping populations appearing higher in the gating hierarchy are excluded from the density. Such a systematic sequential normalization strategy improves upon normalization after dump gates only. 
%
%We combine these tools to develop an integrated data analysis pipeline for ICS assays and use it to analyze two real--world, full size, vaccine trial, ICS data sets from the HVTN, comparing the results against gold standard manual gating.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Immunological endpoints in vaccine trials are measured through a variety of different assays that, often, provide single--cell measurements of multiple, specific intracellular or cell surface proteins, or mRNA expression levels of specific genes in single cells from specific cell sub--populations. Although these measurements are continuous, they are generally discretized for analysis, with each cell called either positive or negative for a measured variable, depending on some predetermined threshold. A motivating example is the intracellular cytokine staining (ICS) flow cytometry assay. This assay is used to assess an individual's immune response to a vaccine by measuring the abundance of antigen--specific T--cell subpopulations producing one or more specific cytokines. In this way, the assay provides phenotypic and functional information about the immune system. ICS assays are used to test multiple antigens and multiple cytokines across hundreds of individuals in a vaccine study. The threshold for calling an individual cell as positive or negative for a given cytokine is predetermined by the gating step of the analysis. The thresholded data is discretized and treated as count data of the number of cytokine positive and negative cells in each stimulated and control sample. The goal of the analysis is to identify the ``vaccine responders''. These are individuals whose immune system produces significantly more cytokine positive cells in response to antigen stimulation than at baseline (the unstimulated control). The rarity of these cell populations makes maximizing the sensitivity and specificity of the assay a primary concern. The typical approach to analysis using Fisher's exact test is problematic for two reasons. It can be overly conservative in detecting true differences for small counts, and the data generally do not meet the assumptions of the test. Specifically, the total cell counts across conditions are generally not fixed since these are generated by independent experiments.  In this paper we present a cohort framework based on an empirical-Bayes mixture of Beta--binomial or Dirichlet--Multinomial distributions, for analyzing count data derived from such thresholded single--cell assays. In general, any single--cell assay in which continuous data can be thresholded to classify individual cells into either ``positive'' or ``negative'' categories with respect to one or more variables and compared across different conditions, is suitable for analysis using our method and the extensions presented herein. Using the ICS assay as a motivating example, our method models cytokine–specific T-cell response across all individuals simultaneously while modelling the stimulated and unstimulated cell counts independently. Using simulations and real--world vaccine trial data, we show that our model increases the sensitivity and specificity for positivity calls in ICS assays compared to Fisher's exact test.
\end{abstract}
\section{Introduction}
Single--cell assays are an important tool in vaccine trials to measure different characteristics of specific immune cell subpopulations. These assays typically measure multiple variables simultaneously for each individual cell in a homogeneous mixture such as whole blood. These variables can be used to classify individual cells in the mixture into more homogeneous phenotypes and functional categories. These, in turn, provide a functional read--out of the immune system. A motivating example is the intracellular cytokine staining assay from flow cytometry, which, in the context of vaccine studies, is used to provide a sensitive and quantitative assessment of antigen--specific T--cell response to a vaccine. It provides both phenotypic information about cell subpopulations (CD4 and CD8 T--cells) in whole blood, and functional information (cytokine expression) about how those subpopulations express different cytokines in response to an antigen stimulus~\cite{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. Assessing a broad T cell response to a vaccine is particularly important in HIV vaccine trials, where the search for immune correlates of protection against HIV progression and infection is ongoing~\cite{Plotkin:2010ve,Horton:2007tsa,Kim:2010fw}.
